
---
title: "Single-Cell RNA-seq Analysis of 10X PBMC Data"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# Introduction

This notebook demonstrates a comprehensive single-cell RNA-seq analysis pipeline using 10X Genomics PBMC data. We'll identify cell types, perform quality control, clustering, and differential expression analysis using the Seurat package in R.

## Objectives

- Load and preprocess 10X PBMC single-cell data
- Perform quality control and filtering
- Identify cell clusters and annotate cell types
- Conduct differential expression analysis
- Generate publication-ready visualizations

# Setup and Installation

```{r install-packages, eval=FALSE}
# Install required packages (run this once)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install Bioconductor packages
BiocManager::install(c("SingleCellExperiment", "scater", "scran"))

# Install CRAN packages
install.packages(c("Seurat", "dplyr", "ggplot2", "patchwork", "RColorBrewer", 
                   "pheatmap", "viridis", "Matrix", "readr", "stringr"))
```

```{r load-libraries}
# Load required libraries
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
library(Matrix)
library(RColorBrewer)
library(pheatmap)
library(viridis)

# Set working directory for Google Colab
if (Sys.getenv("COLAB_RELEASE_TAG") != "") {
  setwd("/content")
}
```

# Data Download and Loading

```{r download-data, eval=FALSE}
# Download 10X PBMC data (3k PBMCs from a healthy donor)
system("wget https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz")
system("tar -xzf pbmc3k_filtered_gene_bc_matrices.tar.gz")
```

```{r load-data}
# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = "filtered_gene_bc_matrices/hg19/")

# Initialize the Seurat object
pbmc <- CreateSeuratObject(
  counts = pbmc.data, 
  project = "pbmc3k", 
  min.cells = 3, 
  min.features = 200
)

# Display basic information
cat("Number of cells:", ncol(pbmc), "\n")
cat("Number of genes:", nrow(pbmc), "\n")
```

# Quality Control Analysis

Quality control is crucial for identifying and filtering out low-quality cells that could affect downstream analysis.

```{r qc-metrics}
# Calculate mitochondrial gene percentage
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# Calculate ribosomal gene percentage
pbmc[["percent.rb"]] <- PercentageFeatureSet(pbmc, pattern = "^RP[SL]")

# Display QC statistics
cat("QC Metrics Summary:\n")
cat("Features per cell - Min:", min(pbmc$nFeature_RNA), "Max:", max(pbmc$nFeature_RNA), "Median:", median(pbmc$nFeature_RNA), "\n")
cat("UMIs per cell - Min:", min(pbmc$nCount_RNA), "Max:", max(pbmc$nCount_RNA), "Median:", median(pbmc$nCount_RNA), "\n")
cat("Mitochondrial % - Min:", round(min(pbmc$percent.mt), 2), "Max:", round(max(pbmc$percent.mt), 2), "Median:", round(median(pbmc$percent.mt), 2), "\n")
```

## QC Visualizations

```{r qc-plots, fig.width=12, fig.height=8}
# Violin plots for QC metrics
p1 <- VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0.1)

# Feature-feature relationships
p2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  geom_hline(yintercept = 5, color = "red", linetype = "dashed")
p3 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  geom_hline(yintercept = c(200, 2500), color = "red", linetype = "dashed")

# Display plots
print(p1)
print(p2 + p3)
```

# Cell Filtering

We filter cells based on the following criteria:
- **Feature count**: 200-2500 genes per cell
- **Mitochondrial content**: <5% mitochondrial genes

```{r cell-filtering}
# Apply filtering criteria
pbmc_filtered <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)

cat("Cells before filtering:", ncol(pbmc), "\n")
cat("Cells after filtering:", ncol(pbmc_filtered), "\n")
cat("Cells removed:", ncol(pbmc) - ncol(pbmc_filtered), "(", round((ncol(pbmc) - ncol(pbmc_filtered))/ncol(pbmc)*100, 2), "%)\n")

# Update pbmc object
pbmc <- pbmc_filtered
```

# Data Normalization

```{r normalization}
# Normalize the data using LogNormalize method
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
cat("Data normalized using LogNormalize method\n")
```

# Highly Variable Features

Identifying highly variable genes helps focus on the most informative features for downstream analysis.

```{r variable-features, fig.width=12, fig.height=6}
# Find highly variable features
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the top 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)
cat("Top 10 highly variable genes:\n")
print(top10)

# Plot variable features with and without labels
p1 <- VariableFeaturePlot(pbmc)
p2 <- LabelPoints(plot = p1, points = top10, repel = TRUE)
print(p1 + p2)
```

# Data Scaling and PCA

```{r scaling-pca}
# Scale the data
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)

# Perform linear dimensional reduction (PCA)
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))

# Examine PCA results
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```

## PCA Visualizations

```{r pca-plots, fig.width=12, fig.height=8}
# PCA loadings and scatter plot
p1 <- VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
p2 <- DimPlot(pbmc, reduction = "pca")

print(p1 / p2)

# PCA heatmap
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```

# Determining Optimal Dimensionality

```{r dimensionality, fig.width=12, fig.height=6}
# JackStraw analysis
pbmc <- JackStraw(pbmc, num.replicate = 100)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20)

p1 <- JackStrawPlot(pbmc, dims = 1:15)
p2 <- ElbowPlot(pbmc)

print(p1 + p2)
```

# Cell Clustering

```{r clustering}
# Find neighbors and clusters
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)

cat("Number of clusters identified:", length(levels(Idents(pbmc))), "\n")
cat("Cluster sizes:\n")
print(table(Idents(pbmc)))
```

# UMAP Visualization

```{r umap, fig.width=10, fig.height=8}
# Run UMAP
pbmc <- RunUMAP(pbmc, dims = 1:10)

# Plot UMAP with cluster labels
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.8) +
  ggtitle("UMAP Clustering of PBMC Cells")
```

# Biomarker Discovery

## Find All Cluster Markers

```{r find-markers}
# Find markers for every cluster
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

# Display top 2 markers per cluster
top2_markers <- pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC)

print(top2_markers)
```

## Cluster-Specific Analysis

```{r cluster-specific}
# Find markers for a specific cluster (cluster 2)
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2, min.pct = 0.25)
cat("Top 5 markers for cluster 2:\n")
print(head(cluster2.markers, n = 5))
```

## Marker Gene Visualization

```{r marker-visualization, fig.width=12, fig.height=10}
# Violin plots for key markers
p1 <- VlnPlot(pbmc, features = c("MS4A1", "CD79A", "CD3E", "CD14"), ncol = 2)

# Feature plots on UMAP
key_features <- c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP", "CD8A")
p2 <- FeaturePlot(pbmc, features = key_features, ncol = 3)

print(p1)
print(p2)
```

## Heatmap of Top Markers

```{r marker-heatmap, fig.width=12, fig.height=10}
# Create heatmap of top 5 markers per cluster
top5_markers <- pbmc.markers %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC)

DoHeatmap(pbmc, features = top5_markers$gene, size = 3) + 
  NoLegend() +
  ggtitle("Top 5 Markers per Cluster")
```

# Cell Type Annotation

Based on canonical marker expression, we can annotate cell types:

```{r cell-annotation}
# Define cell type identities based on markers
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", 
                     "FCGR3A+ Mono", "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)

# Plot with cell type labels
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.8, label.size = 4) + 
  NoLegend() +
  ggtitle("PBMC Cell Types")
```

# Cell Type Distribution Analysis

```{r cell-distribution, fig.width=10, fig.height=6}
# Calculate cell type proportions
cell_counts <- table(Idents(pbmc))
cell_proportions <- prop.table(cell_counts) * 100

cat("Cell Type Distribution:\n")
print(round(cell_proportions, 2))

# Create visualization
cell_type_df <- data.frame(
  CellType = names(cell_proportions),
  Percentage = as.numeric(cell_proportions),
  Count = as.numeric(cell_counts)
)

ggplot(cell_type_df, aes(x = reorder(CellType, -Percentage), y = Percentage)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            vjust = -0.5, size = 3) +
  theme_minimal() +
  labs(title = "Cell Type Distribution in PBMC Sample",
       x = "Cell Type",
       y = "Percentage (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Differential Expression Analysis

## B Cells vs All Other Cells

```{r de-analysis-b-cells}
# Compare B cells vs all other cells
b_cell_markers <- FindMarkers(pbmc, ident.1 = "B", min.pct = 0.25, logfc.threshold = 0.25)
cat("Top 10 B cell markers:\n")
print(head(b_cell_markers, n = 10))
```

## Monocyte Subtypes Comparison

```{r de-analysis-monos}
# Compare CD14+ vs FCGR3A+ Monocytes
mono_markers <- FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = "FCGR3A+ Mono", min.pct = 0.25)
cat("Top 10 markers distinguishing monocyte subtypes:\n")
print(head(mono_markers, n = 10))
```

# Summary and Results

```{r summary}
# Analysis summary
cat("\n=== ANALYSIS SUMMARY ===\n")
cat("Total cells analyzed:", ncol(pbmc), "\n")
cat("Total genes:", nrow(pbmc), "\n")
cat("Number of cell types:", length(unique(Idents(pbmc))), "\n")
cat("Cell types identified:", paste(unique(Idents(pbmc)), collapse = ", "), "\n")
cat("\nCell type distribution:\n")
print(cell_type_df)
```

## Save Results

```{r save-results, eval=FALSE}
# Save the final Seurat object
saveRDS(pbmc, file = "pbmc_final_analysis.rds")

# Export cell metadata
cell_metadata <- data.frame(
  Barcode = colnames(pbmc),
  CellType = Idents(pbmc),
  nFeature_RNA = pbmc$nFeature_RNA,
  nCount_RNA = pbmc$nCount_RNA,
  percent_mt = pbmc$percent.mt
)
write.csv(cell_metadata, "cell_metadata.csv", row.names = FALSE)

# Export marker genes
write.csv(pbmc.markers, "cluster_markers.csv", row.names = TRUE)

cat("Analysis complete! Files saved to working directory.\n")
```

# Session Information

```{r session-info}
sessionInfo()
```

---

**Analysis Complete!** This notebook provides a comprehensive single-cell RNA-seq analysis pipeline. The results show distinct cell populations in human PBMCs with clear separation of major immune cell types including T cells, B cells, monocytes, NK cells, and dendritic cells.
